Exploring mathematical objects from custom-tailored mathematical universes

### 2020-01-12 paper-filmat.tex
Sorted the unsorted

\noindent
Toposes can be pictured as mathematical universes in which we can do
mathematics in. Most mathematicians spend all their professional life in just a
single topos, the so-called \emph{standard topos}. However, besides the
standard topos, there is a colorful host of alternate toposes which are just as
worthy of mathematical study and in which mathematics plays out slightly
differently.

@@

Viewed through such a lens, a given mathematical object can have different
properties than it has when considered normally. In particular, it can have
better properties for the purposes of specific applications, especially if
the topos is custom-tailored to the object in question. This change of
perspective has been used in mathematical practice and demonstrates that
toposes go much beyond being logicians' testbeds. To give just a taste of what
is possible, through the lens provided by an appropriate topos, any given ring
can look like a field.

We argue that toposes and specifically the change in perspective provided by
toposes are ripe for philosophical analysis. In particular, there are the
following connections with topics in the philosophy of mathematics:
\begin{enumerate}
\item Toposes enrich the realism/anti-realism debate in that they point the
picture that the platonic heaven of mathematical objects is not unique: besides
the standard heaven of the standard topos, we could fathom the alternate
heavens of all other toposes, all embedded in a secord-order heaven.
\item Mathematics is not only about studying mathematical objects, but also
about studying the relations between mathematical objects. The distinct view
on mathematical objects provided by any topos uncovers relations which
otherwise remain hidden.
\item In some cases, a mathematical relation can be expressed quite succinctly
using the language of a specific topos and not so succinctly using the language
of the standard topos. This phenomenon showcases the importance of
\emph{appropriate language}.
\item Toposes provide new impetus to study constructive mathematics and
intuitionistic logic, in particular also to restrict to intuitionistic
logic on the meta level and to consider the idea that the platonic heaven might
be governed by intuitionistic logic.
\end{enumerate}
This note only touches on these topics and invites further research.

@@

Formally, a topos is a certain kind of \emph{category}, containing objects and
morphisms between those objects. For instance, the standard topos contains all
sets as its objects and all maps between sets as morphisms. 
% The term \emph{standard topos} reflects that all objects of
% mathematical study can be encoded as sets -- a group is a \emph{set} of
% elements together ...
% \begin{ex}The \emph{standard topos}~$\Set$ contains all
% sets as its objects and all maps between sets as morphisms.\end{ex}
A careful definition of the notion of a topos requires some amount
of category theory, but, as will be outlined in the following sections,
exploring the mathematical universe of a given topos does not.

Put briefly, the axioms state that a topos should share several categorical
properties with the category of sets (namely having finite limits, being
cartesian closed and having a subobject classifier); they ensure that each topos contains its
own versions of familiar mathematical objects such as natural numbers, real
numbers, groups and manifolds, and is closed under the usual constructions
such as cartesian products or quotients.

@@

In the definition of~$\E \models \varphi$, the statement~$\varphi$ can be any
statement in the language of a general version of higher-order predicate
calculus with dependent types. In practice almost any mathematical statement
can be interpreted in a given topos.\footnote{The main exceptions are
statements from set theory, which typically make substantial use of a global
membership predicate~``$\in$''. Toposes only support a typed \emph{local}
membership predicate, where we may write~``$x \in A$'' only in the context of
some fixed sort type~$B$ such that~$x$ is of type~$B$ and~$A$ is of
type~$P(B)$, the power type of~$B$.}

@@

\subsection{The logic of toposes} By their definition as special kinds of
categories, toposes are merely algebraic structures not unlike groups and
vector spaces are. Hence we need to argue why we picture toposes as
mathematical universes while we do not elevate other kinds of algebraic structures
in the same way. For us, the reason is given by the following metatheorem:

\begin{thm}Let~$\E$ be a topos and~$\varphi$ be a statement such that~$\E \models
\varphi$. If~$\varphi$ intuitionistically entails a further statement~$\psi$,
then~$\E \models \psi$.\end{thm}

XXX explain intuitionistically

XXX explain that this metatheorem allows us to \emph{reason} in a topos

XXX include something like this: A natural question is this: \emph{Which of
the familiar mathematical facts of the standard topos carry over to arbitrary
toposes~$\E$?} For any given mathematical statement~$\varphi$ and topos~$\E$,
we can unroll the definition of~$\E \models \varphi$ to try to check
whether~$\varphi$ holds in~$\E$ on an individual case-by-case basis; but it
turns out that there is a general metatheorem ...

XXX Mathematicians are familiar with the fact that the usual objects of
mathematical study are governed by the laws of ordinary \emph{classical
reasoning}. ...

\subsection{Models of set theory}

Toposes can be pictured as ``universes in which we can do mathematics in'' in
much the same way as models of set theory can be viewed in this way. In fact,
to any model~$M$ of a set theory such as~\textsc{zf} or~\textsc{zfc}, there
is a topos~$\Set_M$ such that a statement holds in the internal language
of~$\Set_M$ if and only if it holds in~$M$.\footnote{The topos~$\Set_M$ can be
described as follows: Its objects are the elements of~$M$, that is the things
which~$M$ believes to be sets, and its morphisms are those things which~$M$
believes to be maps. The topos~$\Set_M$ validates the axioms of
$\mathbf{ETCS}$~\cite{XXX}, and for models which are not elementarily
equivalent, their associated toposes will not be equivalent.} This
includes~$V$, the trivial inner model consisting of all sets there are, in case
such a thing exists in one's chosen ontology; in this case the associated
topos~$\Set_V$ is the category of all sets, and a statement holds in~$\Set_V$
if and only if it's true in the ordinary mathematical sense.

Carrying the analogy further, just as using forcing and other techniques we can
construct new models of set theory from given ones, thereby exploring the
set-theoretic multiverse, we can construct new toposes from given toposes.
However, there are two important differences between the notion of mathematical
universe as provided by toposes and as provided by models of set theory, both
regarding the subject matter and the reasons for why we are interested in them.

Firstly, toposes are more general than models of set theory. By definition, a
model of~$\mathbf{ZFC}$ will always satisfy the axioms of~$\mathbf{ZFC}$; in
contrast, most toposes do not even validate the law of excluded middle, much
less so the axiom of choice (which implies the law of excluded middle in
presence of the remaining axioms of intuitionistic Zermelo--Fraenkel set
theory~\cite{XXX}). This failure of classicality is often for interesting
geometric or algebraic reasons instead of logical issues, as will be
detailed in Section~\ref{several-toposes}.

Secondly, there is a difference in motivation. The main philosophical reason
for studying models of set theory is to study which notions of sets are
coherent: Does the cardinality of the reals need to be the cardinal directly
succeeding~$\aleph_0$, the cardinality of the naturals? No, there are models of
set theory in which the continuum hypothesis fails. Do non-measurable sets of
reals need to exist? No, in models of~$\mathbf{ZF}+\mathbf{AD}$,
Zermelo--Fraenkel set theory plus the axiom of determinacy, it's a theorem that
every subset of~$\RR^n$ is Lebesgue-measurable. Can the axiom of choice be
added to the axioms of~$\mathbf{ZF}$ without causing inconsistency? Yes, if~$M$
is a model of~$\mathbf{ZF}$ then~$L^M$, the set of definable sets of~$M$, is a
model of~$\mathbf{ZFC}$.~\cite{XXX, maybe SEP?}

While toposes can be used for similar such purposes, and indeed have been,
especially to explore the various intuitionistic notions of sets, an important
aspect of topos theory is that toposes are used to explore the standard
mathematical universe: Truth in the \effective topos tells us what is
computable; truth in sheaf toposes tells us what's true locally, in a fashion
which depends continuously on parameters; the little Zariski topos of a general
commutative ring can be used to pretend that the given ring is a field; toposes
adapted to synthetic differential geometry can be used to rigorously work with
infinitesimals. All of these examples will be presented in more detail in
Section~\ref{several-toposes}.

In a certain precise way, toposes allow us to study the common objects of
mathematics from a different point of view -- one such view for every topos --
and it is a beautiful and intriguing fact that with the sole exception of the
law of excluded middle, the laws of logic apply to mathematical objects also
when viewed through the lens of a specific topos.

>>>>

\noindent
Toposes can be pictured as mathematical universes in which we can do
mathematics. Most mathematicians spend all their professional life in just a
single topos, the so-called \emph{standard topos}. However, besides the
standard topos, there is a colorful host of alternate toposes which are just as
worthy of mathematical study and in which mathematics plays out slightly
differently.

@@

Viewed through such a lens, a given mathematical object can have different
properties than when considered normally. In particular, it can have
better properties for the purposes of specific applications, especially if
the topos is custom-tailored to the object in question. This change of
perspective has been used in mathematical practice and demonstrates that
toposes go much beyond being logicians' testbeds. To give just a taste of what
is possible, through the lens provided by an appropriate topos, any given ring
can look like a field and hence mathematical techniques for fields also apply,
through the lens, to rings.

We argue that toposes and specifically the change in perspective provided by
toposes are ripe for philosophical analysis. In particular, there are the
following connections with topics in the philosophy of mathematics:
\begin{enumerate}
\item Toposes enrich the realism/anti-realism debate in that they paint the
picture that the platonic heaven of mathematical objects is not unique: besides
the standard heaven of the standard topos, we could fathom the alternate
heavens of all other toposes, all embedded in a secord-order heaven.
\item Mathematics is not only about studying mathematical objects, but also
about studying the relations between mathematical objects. The distinct view
on mathematical objects provided by any topos uncovers relations which
otherwise remain hidden.
% XXX: mention Olivia?
\item In some cases, a mathematical relation can be expressed quite succinctly
using the language of a specific topos and not so succinctly using the language
of the standard topos. This phenomenon showcases the importance of
\emph{appropriate language}.
\item Toposes provide new impetus to study constructive mathematics and
intuitionistic logic, in particular also to restrict to intuitionistic
logic on the meta level and to consider the idea that the platonic heaven might
be governed by intuitionistic logic.
\end{enumerate}
This note touches on these topics and invites further research.

@@

Formally, a topos is a certain kind of \emph{category}, containing objects and
morphisms between those objects. The formal definition, recorded here only for
reference, requires some amount of category theory, but, as will be outlined in
the following sections, exploring the mathematical universe of a given topos
does not.

\begin{defn}A \emph{topos} (more precisely \emph{elementary topos} or
\emph{logos} with a natural numbers object) is a category which has all finite
limits, is cartesian closed and has a subobject classifier.\end{defn}

Put briefly, the axioms state that a topos should share several categorical
properties with the category of sets; they ensure that each topos contains its
own versions of familiar mathematical objects such as natural numbers, real
numbers, groups and manifolds, and is closed under the usual constructions
such as cartesian products or quotients.
The prototypical topos is the standard topos:

\begin{ex}The \emph{standard topos}~$\Set$ contains all
sets as its objects and all maps between sets as morphisms.\end{ex}

@@

In the definition of~$\E \models \varphi$, the statement~$\varphi$ can be any
statement in the language of a general version of higher-order predicate
calculus with dependent types. In practice almost any mathematical statement
can be interpreted in a given topos.\footnote{The main exceptions are
statements from set theory, which typically make substantial use of a global
membership predicate~``$\in$''. Toposes only support a typed \emph{local}
membership predicate, where we may write~``$x \in A$'' only in the context of
some fixed type~$B$ such that~$x$ is of type~$B$ and~$A$ is of
type~$P(B)$, the power type of~$B$.}

@@

\subsection{The logic of toposes} By their definition as special kinds of
categories, toposes are merely algebraic structures not unlike groups or vector
spaces. Hence we need to argue why we picture toposes as mathematical universes
while we do not elevate other kinds of algebraic structures in the same way.
For us, this usage is justified by the following metatheorem:

\begin{thm}\label{thm:reasoning}Let~$\E$ be a topos and let~$\varphi$ be a
statement such that~$\E \models \varphi$. If~$\varphi$ intuitionistically
entails a further statement~$\psi$ (that is, if it is provable in
intuitionistic logic that~$\varphi$ entails~$\psi$), then~$\E \models
\psi$.\end{thm}

This metatheorem allows us to \emph{reason} in toposes. When first exploring a
new topos~$\E$, we need to employ the Kripke--Joyal translation rules each time
we want to check whether a statement holds in~$\E$. But as soon as we
have amassed a stock of statements known to be true in~$\E$, we can find more
by deducing their logical consequences.

For instance, in any topos where the statement ``any map~$\RR \to \RR$ is
continuous'' is true, also the statement ``any map~$\RR \to \RR^2$ is
continuous'' is, since there is an intuitionistic proof that a map into a
higher-dimensional Euclidean space is continuous if its individual components
are.

The only caveat of Theorem~\ref{thm:reasoning} is that toposes generally only
support intuitionistic reasoning, not the full power of the ordinary
\emph{classical reasoning}. That is, within most toposes, the law of excluded
middle ($\varphi \vee \neg\varphi$) and the law of double negation elimination
($\neg\neg\varphi \Rightarrow \varphi$) are not available.

While it may appear that these two laws pervade any mathematical theory, in
fact a substantial amount of mathematics can be developed intuitionistically
(see for
instance~\cite{mines-richman-ruitenburg:constructive-algebra,lombardi-quitte:constructive-algebra}
for constructive algebra,~\cite{bishop-bridges:constructive-analysis} for
constructive analysis
and~\cite{bauer:int-mathematics,bauer:video,melikhov:intuitionistic-logic} for
further references) and hence the alternate universes provided by toposes
cannot be too strange: In any topos, there are infinitely many prime numbers,
the square root of two is not rational and the powerset of the naturals is
uncountable.

\begin{defn}A topos~$\E$ is \emph{boolean} if and only if the laws of classical
logic are true in~$\E$.\end{defn}

The standard topos is boolean if and only if, as is commonly supposed, the laws
of classical logic hold on the meta level. Most toposes of interest are not
boolean, irrespective of one's philosophical commitments about the meta level,
and conversely some toposes are boolean even if classical logic is not
available on the meta level.

\begin{rem}The axiom of choice (which is strictly speaking not part of
classical logic, but of classical set theory) is also not available in most
toposes. By \emph{Diaconescu's theorem}, the axiom of choice implies the law of
excluded middle in presence of other axioms which are available in any topos.
\end{rem}

%A natural question is this: \emph{Which of
%the familiar mathematical facts of the standard topos carry over to arbitrary
%toposes?} For any given mathematical statement~$\varphi$ and topos~$\E$,
%we can unroll the definition of~$\E \models \varphi$ to try to check
%whether~$\varphi$ holds in~$\E$ on an individual case-by-case basis; but since
%any topos supports intuitionistic reasoning, we may at once conclude: \emph{Any
%theorem of constructive mathematics, that is any theorem deduced using only
%intuitionistic logic, is valid in any topos.}

%Mathematicians are familiar with the fact that the usual objects of
%mathematical study are governed by the laws of ordinary \emph{classical
%reasoning}. ...

\subsection{Relation to models of set theory} In set theory, philosophy and
logic, models of set theories ... XXX

Toposes can be pictured as
``universes in which we can do mathematics'' in much the same way as models of
set theory. Just as using forcing and other techniques we can construct new
models of set theory from given ones, thereby exploring the set-theoretic
multiverse, we can construct new toposes from given toposes.

However, there are important differences between the notion of mathematical
universe as provided by toposes and as provided by models of set theory, both
regarding the subject matter and the reasons for why we are interested in them.

Firstly, toposes are more general than models of set theory. By definition, a
model of \textsc{zfc} will always satisfy the axioms of \textsc{zfc}; in
contrast, most toposes do not even validate the law of excluded middle, much
less so the axiom of choice.

Secondly, there is a shift in emphasis. An important philosophical objective
for studying models of set theory is to explore which notions of sets are
coherent: Does the cardinality of the reals need to be the cardinal directly
succeeding~$\aleph_0$, the cardinality of the naturals? No, there are models of
set theory in which the continuum hypothesis fails. Do non-measurable sets of
reals need to exist? No, in models of~$\textsc{zf}+\textsc{ad}$,
Zermelo--Fraenkel set theory plus the axiom of determinacy, it is a theorem that
every subset of~$\RR^n$ is Lebesgue-measurable. Can the axiom of choice be
added to the axioms of~\textsc{zf} without causing inconsistency? Yes, if~$M$
is a model of~\textsc{zf} then~$L^M$, the set of definable sets of~$M$, is a
model of~\textsc{zfc}.  %~\cite{XXX, maybe SEP?}

While toposes can be used for similar such purposes, and indeed have been,
especially to explore the various intuitionistic notions of sets, an important
aspect of topos theory is that toposes are used to explore the standard
mathematical universe: truth in the \effective topos tells us what is
computable; truth in sheaf toposes tells us what's true locally; toposes
adapted to synthetic differential geometry can be used to rigorously work with
infinitesimals. All of these examples will be presented in more detail in the
next sections.

In a sense which can be made precise, toposes allow us to study the usual
objects of mathematics from a different point of view -- one such view for
every topos -- and it is a beautiful and intriguing fact that with the sole
exception of the law of excluded middle, the laws of logic apply to
mathematical objects also when viewed through the lens of a specific topos.

In fact,
to any model~$M$ of a set theory such as~\textsc{zf} or~\textsc{zfc}, there
is a topos~$\Set_M$ such that a statement holds in the internal language
of~$\Set_M$ if and only if it holds in~$M$.\footnote{The topos~$\Set_M$ can be
described as follows: Its objects are the elements of~$M$, that is the things
which~$M$ believes to be sets, and its morphisms are those things which~$M$
believes to be maps. The topos~$\Set_M$ validates the axioms of
$\mathbf{ETCS}$~\cite{XXX}, and for models which are not elementarily
equivalent, their associated toposes will not be equivalent.} This
includes~$V$, the trivial inner model consisting of all sets there are, in case
such a thing exists in one's chosen ontology; in this case the associated
topos~$\Set_V$ is the category of all sets, and a statement holds in~$\Set_V$
if and only if it's true in the ordinary mathematical sense.
